Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-31 response-to-reviewers.tex

\newenvironment{reply}
   {\medskip \noindent \begin{sf}\textbf{Reply}:\  }
   {\medskip \end{sf}}

\newcommand{\shortreply}[2][]{\medskip \noindent \begin{sf}\textbf{Reply}:\  #2
    \ifthenelse{\equal{#1}{}}{}{ \hfill \footnotesize (#1)}%
    \medskip \end{sf}}

\begin{document}

\section*{Response to the editor}
% General intro text goes here
Thank you for considering this manuscript for publication.  
We have addressed the comments from both reviewers point-by-point
below.

@@

\begin{point}
"There is currently no efficient, FAIR representation...". Just a nit and
feel free to ignore, but the solution you present is technically "current".
\end{point}
\begin{reply}
You are right, but for narrative purposes it is simpler to pose VCF Zarr 
as new, so we'd prefer to keep this as it is.
\end{reply}

\begin{point}
In Figure 2, the zarr line is occluded by the sav line and hard to see.
\end{point}
\begin{reply}
The benchmarks have changed slightly, and the lines are now more separated.
\end{reply}

\reviewersection
\section*{Reviewer 2 comments}

\begin{point}
The paper presents an encoding of the VCF data using Zarr to enable fast
retrieving subsets of the data. A vcf2arr conversion was provided and validated
on both simulated and real-world data sets. The topic of this work is
interesting and of good values, however, the experimental studies and
contributions should be considerable improved.
\end{point}
\begin{reply}
Thank you for your valuable feedback; it has resulted in a much improved
manuscript. We have significantly extended the scope 
of this work by providing 
\begin{itemize}
\item The vcztools program, which converts from Zarr to VCF efficiently.
\item A new case study on the Our Future Health genotype data, for 
651K samples.
\item A new case study on the All of Us exome-like data, for 245K
samples.
\item A new case study on Norway Spruce data, for 1063 samples
and 3.75 \emph{billion} variants.
\item A new case study on SARS-CoV-2 data for 4.3 million samples.
\item A new section demonstrating a propotype of the SAIGE software
incorporating support for VCF Zarr, with excellent performance.
\item A new section exploring the use of VCF Zarr in cloud computing 
platforms, showing how data processing rates of up to 25.5GiB/s
are possible.
\item A new section exploring the use of GPU acceleration on VCF Zarr data.
\end{itemize}
We hope that you agree that these extensive additions on 
boundary-pushing datasets, demonstrate the 
real-world value and transformative potential of VCF Zarr.
\end{reply}

\begin{point}
The proposed method is simply a conversion from VCF to Zarr format. Since
both are existing formats, the contributions and originality of this work are
not impressive.
\end{point}
\begin{reply}
We believe that the extensive work detailed in the previous response
addresses this criticism.
\end{reply}

\begin{point}
The compression and query performance is the main concern of this work. The
method should be compared with other state-of-the-art queriable VCF compressors
like GTC, GBC, and GSC. [References ommitted]
\end{point}
\begin{reply}
We disagree on this point, and respectfully point out that for our 
purposes, these methods are essentially equivalent to Genozip
(which we have included in our benchmarks as a representative 
of this class of method). 
As we argue in the ``Calculating with the genotype matrix'' section,
a fundamental point
is that compression methods that only provide access to data 
output via VCF text are limiting the rate at which the data can be 
processed to the rate at which VCF text can be parsed. Producing VCF text
as output cannot be seen as the end point of analysis, and the rate at which it can be
done (beyond a certain minimum value of around 120 MiB/s, as we argue in the 
new vcztools section) is of no real practical interest. Our benchmarks 
in e.g. Figures 3 and 4 show that the Genozip like approach of 
of highly compressed genotype data that is accessed only via VCF text
cannot form the basis of efficient processing pipelines. It is 
intrinsically and irredeemably inefficient.

We have included references to GTC, GBC and GSC, and discussing them in the 
``Storing genetic variation data'' section, directing the reader to 
the benchmarks performed  in the ``Calculating with the genotype matrix''
section.
\end{reply}

\begin{point}
The method should be evaluated on more real VCF data sets.
\end{point}
\begin{reply}
We have added four new case studies on large scale real datasets
on different organisms with entirely different genome properties.
\end{reply}

### 2025-01-31 paper.tex

One of the key benefits of Zarr is its cloud-native design,
but it also works well on standard file systems, where
arrays and chunks are stored hierarchically in directories
and files (storage as a single Zip archive is also supported,
and has certain advantages as discussed later).
To enable comparison with the existing file-based ecosystem
of tools, we focus on Zarr's file system chunk storage in a series of illustrative
benchmarks in the following sections before presenting cloud-based benchmarks.
We compare primarily with
VCF/BCF based workflows using \texttt{bcftools} because this
is the standard practice, used in the vast majority of cases.
We also compare with two representative recent specialised utilities;
see~\cite{danek2018gtc,zhang2023gbc} for further benchmarks of
these and other tools.
Genozip~\cite{lan2020genozip,lan2021genozip} is a tool focused
on compression performance, which uses a custom file format
and a CLI to extract VCF as text with various filtering options.
Savvy~\cite{lefaive2021sparse} is an extension of BCF which
takes advantage of sparsity in the genotype matrix as well
as using PBWT-based approaches for improved compression.
Savvy provides a CLI as well as a C++ API.
Our benchmarks are based on genotype data
from subsets of a large and highly realistic
simulation of French-Canadians~\cite{anderson2023on}
(see Methods for details on the dataset and benchmarking methodology).
Note that while simulations cannot capture
all the subtleties of real data, the allele frequency
and population structure patterns in this dataset
have been shown to closely follow
observations~\cite{anderson2023on} and so it provides
a reasonable and easily reproducible data point
when comparing such methods.
The simulations only contain genotypes without any additional
high-entropy QC fields, which is unrealistic
(see the later case-studies
for benchmarks on datasets that contain a wide variety
of such fields).
Note, however, that such minimal, genotype-only data
is something of a best-case scenario for specialised genotype
compression methods using row-wise storage.

@@

\subsection{Calculating with the genotype matrix}
Storing genetic variation data compactly is important, but it is also
important that we can analyse the data efficiently. Bioinformatics
workflows tend to emphasise text files and command line utilities
that consume and produce text~\citep[e.g.][]{buffalo2015bioinformatics}.
Thus, many tools that compress VCF data provide a command line
utility with a query language to restrict the records
examined, perform some pre-specified calculations and finally
output some text, typically VCF or tab/comma separated
values~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
zhang2023gbc}. %GBC
These pre-defined calculations are by necessity limited in scope, however,
and the volumes of text involved in Biobank scale datasets
make the classical approach of custom
analyses via Unix utilities in pipelines prohibitively slow. Thus,
methods have begun to provide Application Programming Interfaces
(APIs), providing efficient access to genotype and other VCF
data~\cite[e.g.][]{kelleher2013processing,lefaive2021sparse,
wertenbroek2022xsi}. By providing programmatic access,
the data can be retrieved from storage, decoded and then analysed
in the same memory space without additional copies and
inter-process communication through pipes.

>>>>

One of the key benefits of Zarr is its cloud-native design,
but it also works well on standard file systems, where
arrays and chunks are stored hierarchically in directories
and files (storage as a single Zip archive is also supported,
and has certain advantages as discussed later).
To enable comparison with the existing file-based ecosystem
of tools, we focus on Zarr's file system chunk storage in a series of illustrative
benchmarks in the following sections before presenting cloud-based benchmarks.
We compare primarily with
VCF/BCF based workflows using \texttt{bcftools} because this
is the standard practice, used in the vast majority of cases.
We also compare with two representative recent specialised utilities;
see~\cite{danek2018gtc,zhang2023gbc,luo2024gsc} for further benchmarks of
these and other tools.
Genozip~\cite{lan2020genozip,lan2021genozip} is a tool focused
on compression performance, which uses a custom file format
and a CLI to extract VCF as text with various filtering options.
Savvy~\cite{lefaive2021sparse} is an extension of BCF which
takes advantage of sparsity in the genotype matrix as well
as using PBWT-based approaches for improved compression.
Savvy provides a CLI as well as a C++ API.
Our benchmarks are based on genotype data
from subsets of a large and highly realistic
simulation of French-Canadians~\cite{anderson2023on}
(see Methods for details on the dataset and benchmarking methodology).
Note that while simulations cannot capture
all the subtleties of real data, the allele frequency
and population structure patterns in this dataset
have been shown to closely follow
observations~\cite{anderson2023on} and so it provides
a reasonable and easily reproducible data point
when comparing such methods.
The simulations only contain genotypes without any additional
high-entropy QC fields, which is unrealistic
(see the later case-studies
for benchmarks on datasets that contain a wide variety
of such fields).
Note, however, that such minimal, genotype-only data
is something of a best-case scenario for specialised genotype
compression methods using row-wise storage.

@@

\subsection{Calculating with the genotype matrix}
Storing genetic variation data compactly is important, but it is also
important that we can analyse the data efficiently. Bioinformatics
workflows tend to emphasise text files and command line utilities
that consume and produce text~\citep[e.g.][]{buffalo2015bioinformatics}.
Thus, many tools that compress VCF data provide a command line
utility with a query language to restrict the records
examined, perform some pre-specified calculations and finally
output some text, typically VCF or tab/comma separated
values~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
zhang2023gbc,%GBC
luo2024gsc}. %GSC
These pre-defined calculations are by necessity limited in scope, however,
and the volumes of text involved in Biobank scale datasets
make the classical approach of custom
analyses via Unix utilities in pipelines prohibitively slow. Thus,
methods have begun to provide Application Programming Interfaces
(APIs), providing efficient access to genotype and other VCF
data~\cite[e.g.][]{kelleher2013processing,lefaive2021sparse,
wertenbroek2022xsi}. By providing programmatic access,
the data can be retrieved from storage, decoded and then analysed
in the same memory space without additional copies and
inter-process communication through pipes.
